---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# StarWarsStats

<!-- badges: start -->

<!-- badges: end -->

The goal of this R Package is to provide a set of utility functions to easily load, filter, validate, and analyze data from the 'dplyr::starwars' data set, focusing on character attributes, names, heights, genders, and associated starships. The package offers robust, well documented examples of R programming techniques, including, 'dplyr' wrangling, input validation, and error handling within a fun, accessible domain. It was created as an educational exercise to demonstrate practical R function writing and documentation standards, and is primarily intended for students learning R programming who want hands on examples of data manipulation and building reuseable functions within validation checks.

## Installation

Since this package is not yet on CRAN, you can install the development version directly from [GitHub](https://github.com/skywalshUSF/StarWarsStats) using the 'remotes' package.

### Prerequisites

You must have the following packages installed:

remotes: To install packages directly from GitHub.

dplyr: This package contains the 'starwars' dataset and the functions the package relies on.

You can install these prerequisites by running the following command in your R console:

```{r}
# install.packages(c("remotes", "dplyr"))
```


### Install the 'StarWarsStats' Package

Once the remotes package is installed, use the install_github() function:

```{r}
# Install 'StarWarsStats' package

# remotes::install_github("skywalshUSF/StarWarsStats")
```

### Load the Package

After installation is complete, you can load the package and start using your functions:

```{r}
# library(swstats)
```

## Data

The 'StarWarsStats' package uses the widely available 'starwars' dataset provided by the 'dplyr' package.

Source: The data is a curated selection from the Star Wars API (SWAPI), packaged for use in R.

Content: It contains 87 characters across 14 variables, covering biographical, physical, and historical information (e.g., name, height, species, and films appeared in).

Key Detail (Birth Year): Note that birth years are measured relative to the Battle of Yavin (BBY). A positive number (e.g., 41.9) means the event occurred Before the Battle of Yavin, and a negative number means it occurred After the Battle of Yavin (ABY).

```{r}
# To see a full description of all 14 variables in the dataset, run the following command in your R console:

# ?starwars
# summary(starwars)
```


## Examples

This is a basic example which shows you how to solve a common problem:

```{r example}
## basic example code
```

